---
title: "Assignment - Week 03"
author: "Yara Roth"
date: "2026-03-09"
output: html_document
---
# Code

## Setup
```{r setup}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(Rsubread)
  library(rtracklayer)
  library(Biostrings)
  library(Rfastp)
  library(epiwraps)
})
ah <- AnnotationHub()

library(rstudioapi)
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

## Download ChiP-seq for CTCF protein
```{r download_fastqz}
dir.create("data")
dir.create("data/raw")

if (!file.exists("data/raw/ctcf_input.fastq.gz")) {
  download.file("https://www.encodeproject.org/files/ENCFF127RRR/@@download/ENCFF127RRR.fastq.gz", "data/raw/ctcf_input.fastq.gz", timeout = 5000)
} else {
  print("File already exists.")
}
```

## Quality Control and Trimming

### Quality Control
Now we will perform quality control, to look at the quality of the reads.
I did perform this directly in my terminal.

```{bash engine.opts='-l', eval=FALSE}
mkdir -p data/raw_fastqc
fastqc -o data/raw_fastqc --threads 4 data/raw/*.fastq.gz
```

The html file with the results can be found in ``week03/data/raw_fastqc/``.  
The quality of the reads is not optimal, we have a mean quality score of ~23, with increasing variability towards the end of the reads. No adapter sequences were identified.

### Trimming

```{r qc-trimming}
dir.create("data/rfastp.trimmed")
qc <- Rfastp::rfastp(
  read1="data/raw/ctcf_input.fastq.gz",
  thread=4,
  overrepresentationAnalysis=TRUE,
  adapterTrimming = TRUE,
  outputFastq="data/rfastp.trimmed/ctcf_trimmed"
)
```

Reads get trimmed if mean quality drops below 20 for a bp window of 4 or if an adapter gets detected. 
98.26% passed the filtering, meaning they had sufficient quality (less than 40% bases below Q15) and had a goon N content.

```{r}
Rfastp::curvePlot(qc, curve="content_curves")
```

```{r}
Rfastp::curvePlot(qc, curve="quality_curves")
```

## Alignment
Now we want to get the genome sequence for the drosophila melangaster in a TwoBit format for BDGP6 (we already downloaded it last week).
We will use the latest available genome.

### Retrieving drosophila genome
```{r drosophila_genome}
query_results <- query(ah, c("Drosophila melanogaster","TwoBitFile","dna_sm","BDGP6"))

mcols(query_results)[, c("title", "rdatadateadded")]
query_results[length(query_results)]
genome_d <- ah[["AH106160"]]

dir.create("data/BDGP6_genome")
export(import.2bit(genome_d), "data/BDGP6_genome/genome.fasta.gz", compress=TRUE)
```

### Building Genome Index

Now we will build the genome index based on our BDGP6 genome, which will allow us to map/align the reads.

```{r}
if (!file.exists("data/BDGP6_genome/rsubread.files")) {
  Rsubread::buildindex("data/BDGP6_genome/rsubread", reference="data/BDGP6_genome/genome.fasta.gz")
} else {
  message("Index already built.")
}
```

### Alignment

Now we will perform the mapping.

```{r alignment}
dir.create("data/aligned")
align.stats <- Rsubread::align(
  index="data/BDGP6_genome/rsubread",
  type="dna",
  readfile1="data/rfastp.trimmed/ctcf_trimmed_R1.fastq.gz",output_file="data/aligned/ctcf.bam",
  nthreads=6,
  sortReadsByCoordinates=TRUE
)
align.stats
```

We have a mappability of 91.8% which is very high. The bam file can be found in ``data/aligned/ctcf.bam``.

## Peak calling
```{r MACSr-installation}
BiocManager::install("MACSr")
library(MACSr)
```

Now we will extract the peaks and their supplementary information (location) using: 
```{r peak-calling}
dir.create("data/peaks")
callpeak(
  tfile="data/aligned/ctcf.bam",
  gsize="dm",
  outdir = "data/peaks",
  name = "ctcf"
)

peaks <- rtracklayer::import("data/peaks/ctcf_peaks.narrowPeak", format="narrowPeak")
peaks
```

# Report
## Mappability
3'505'377 out of total 3'816'999 reads were mapped, which corresponds to a mappability of 91.8%.

## Number of identified peaks
A total of 5405 peaks were identified for the protein CTCF. 

## Peak Signal inside a gene
To identify peaks that lie within a gene, we first need to extract the genes from the drosophila EnsDb.

```{r download-drosophila-ensdb}
query_results <- query(ah, c("Drosophila","EnsDb","BDGP6"))
query_results[length(query_results)]
ensdb_d <- ah[["AH119285"]]
```

```{r retireve-genes}
genes_d <- genes(ensdb_d)
head(genes_d)
```

When I tried to extract the peaks that lie within a gene, I got a warning that I have a mismatch between the naming conventions of peaks and genes_d. I tried to solve it with the following chunk. Unfortunately, I still received a warning after and did not manage to resolve the issue completely. Maybe we could look at that in the lecture.
```{r}
seqlevelsStyle(peaks) <- seqlevelsStyle(genes_d)
```

```{r peaks-in-genes}
peaks_in_genes <- suppressWarnings(peaks[overlapsAny(peaks, genes_d, type="within")])
peaks_in_genes
```

Now i will select the biggest peak we have (so the one with the highest score).

```{r peak-selection}
biggest_peak_idx <- which.max(peaks_in_genes$score)
biggest_peak <- peaks_in_genes[biggest_peak_idx]
```

Next I'll extraxt the name of the gene, in which the peak is located.
```{r identify-name-gene}
gene_idx <- suppressWarnings(subjectHits(findOverlaps(biggest_peak, genes_d)))
gene_name <- genes_d[gene_idx]$gene_name
gene_name
```

And lastly we can plot everything:
```{r peak-plot}
plotSignalTracks(
  list(ctcf = "data/aligned/ctcf.bam"),
  region = biggest_peak,
  extend = 500,
  genomeAxis = TRUE,
  ensdb = ensdb_d,
  transcripts = "full",
  main = paste("CTCF peaks inside the", gene_name, "gene")
)
```


