---
title: "Assignment - Week 02"
author: "Yara Roth"
date: "2026-04-04"
output: html_document
---
## Exercise 1: 
### Setup
```{r setup}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(ensembldb)
  library(GenomicRanges)
  library(epiwraps)
})
```

### Mouse ensdb
```{r mouse_ensdb}
ah <- AnnotationHub()
query_results <- query(ah, c("Mus Musculus", "EnsDb","102","GRCm38"))
head(query_results)

ensdb_m <- ah[["AH89211"]]
ensdb_m
```

### Mouse genome sequence
```{r mouse_genome}
query_results <- query(ah, c("Mus Musculus","TwoBitFile","102","GRCm38"))
head(query_results)
```

there are different entries (cdna, dna_rm,dna_sm) -> so we need to specify dna_sm when retireving it:

```{r mouse_genome_dnasm}
query_results <- query(ah, c("Mus Musculus","TwoBitFile","102","GRCm38","dna_sm"))
head(query_results)
genome_m <- ah[["AH88477"]]
```

### Drosophila genome sequence
```{r drosophila_genome}
query_results <- query(ah, c("Drosophila melanogaster","TwoBitFile","dna_sm","BDGP6"))
head(query_results)
```

we again have multiple entires, so we can look at the dates the genome were added (we will pick the latest one)

```{r drosophila_genome_latest}
mcols(query_results)[, c("title", "rdatadateadded")]
query_results[length(query_results)]
genome_d <- ah[["AH106160"]]
```

## Exercise 2:
### Extract the number of (unique) gene IDs and symbols:
I'll first inspect the genes(ensdb_m)
```{r inspect_ensdb_m}
mouse_genes <- genes(ensdb_m)
mcols(mouse_genes)
```

Now we only select the protein coding genes and extract the number of unique gene IDs and gene symbols

```{r pcg_mouse}
mouse_pqg <- genes(ensdb_m, filter=~gene_biotype=="protein_coding")
length(mouse_pqg)

number_gene_id <- length(unique(mouse_pqg$gene_id))
number_symbol <- length(unique(mouse_pqg$symbol))
```

Number of unique gene IDs: `r number_gene_id`  
Number of unique gene symbols `r number_symbol`  
There are less unique gene symbols than gene ids, meaning that some of the symbols are not unique. 

### Plot the distribution of (spliced) transcript length
```{r length transcripts}
# we can group the exons by transcript, meaning each transcript has a list of exons organized together:
mouse_exons <- exonsBy(ensdb_m,filter=~gene_biotype=="protein_coding")

# this shows me the length of the individual exon in a certain transcript: 
mouse_exon_lengths <- width(mouse_exons)

# therefore we can add up the individual exon length, resulting in total transcript length
mouse_transcript_lengths <- sum(width(mouse_exons))
head(mouse_transcript_lengths)
```

```{r histogram}
hist(mouse_transcript_lengths,
     main="Distribution of the length of protein-coding transcripts",
     xlab="Length Transcripts [bp]",
     col="#8DBBDC",
     breaks=50)
```

The histogram looks very broad, so I'll check the maximal length of transcript we have to see whether that is really needed.

```{r}
max(mouse_transcript_lengths)
```

Indeed we have transcripts up to a length of 123179 bp meaning I'll leave the histogram as it is. 