---
title: "Assignment - Week 07"
author: "Yara Roth"
date: "2026-04-17"
output: html_document
---

# Setup

```{r setup}
suppressPackageStartupMessages({
  library(motifmatchr)
  library(MotifDb)
  library(universalmotif)
  library(ggplot2)
  library(sechm)
  library(limma)
  library(betterChromVAR)
  library(epiwraps)
  library(AnnotationHub)
  library(memes)
  library(ensembldb)
  library(magick)
  library(BSgenome.Mmusculus.UCSC.mm10)
})

library(rstudioapi)
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
dir.create("data")

```

# Part I
To look at the accessability of the SP1 TF motif we will do the following:
1. Generate NF and mono-nucleosome tracks from ATAC-seq data  
2. Retrieve TF motif  
3. Find motif sites in genome  
4. Look at ATAC signals around the motif occurences  

## Downloading ATAC-seq data from Chr19
```{r, eval=FALSE}
download.file("https://ethz-ins.org/content/w7/atac.chr19.bam", "data/atac.chr19.bam", mode="wb")
download.file("https://ethz-ins.org/content/w7/atac.chr19.bam.bai", "data/atac.chr19.bam.bai", mode="wb")
```

## Preparing tracks
```{r create-tracks}
bam <- "data/atac.chr19.bam"

# create track to look at insertion profiles of nucleosome-free fragment
epiwraps::bam2bw(bam, output_bw = "data/NF_cuts.bw", paired=TRUE, binWidth=1L, type="ends", minFragLength=30, 
       maxFragLength=120, trim=4L, extend=3L, forceSeqlevelsStyle = "Ensembl")

# create track using the centers of mono-nucleosome fragments
bam2bw(bam, output_bw = "data/mono_centers.bw", paired=TRUE, binWidth=5L, minFragLength=140, trim=4L,
       maxFragLength=220, type="center", extend=10L, forceSeqlevelsStyle = "Ensembl")
```

## Obtaining sites with a SP1 or CTCF motif
I chose to make the analysis for both CTCF and SP1, which allows to compare the footprints of both.

```{r motif-sp1-ctcf}
# we will first start by retrieving the SP1 TF motifs from the MotifDb database

motif_sp1 <- MotifDb::query(MotifDb, c("SP1","Mus"))[[1]]
motif2_sp1 <- convert_motifs(motif_sp1, class="TFBSTools-PFMatrix")

motif_ctcf <- MotifDb::query(MotifDb, c("CTCF","Mus"))[[1]]
motif2_ctcf <- convert_motifs(motif_ctcf, class="TFBSTools-PFMatrix")
```

```{r mouse-genome}
# retrieving latest mouse genome
ah <- AnnotationHub()

query_results <- AnnotationHub::query(ah, c("Mus Musculus","TwoBitFile","101","GRCm38","primary_assembly"))
head(query_results)

genome_m <- ah[["AH84785"]]
chr19 <- import(genome_m)["19"]
```

```{r motif-in-genome}
# find SP1 motif matches across Chr19
moi_sp1 <- motifmatchr::matchMotifs(motif2_sp1, chr19, out="positions", p.cutoff=1e-5)[[1]]
moi_sp1 <- as(setNames(moi_sp1,names(chr19)), "GRanges")

moi_ctcf <- motifmatchr::matchMotifs(motif2_ctcf, chr19, out="positions", p.cutoff=1e-5)[[1]]
moi_ctcf <- as(setNames(moi_ctcf,names(chr19)), "GRanges")
```

## Plotting ATAC-signals around motifs
### ATAC signals for CTCF motifs
```{r, fig.width=8, fig.height=4}
tracks <- list.files("data/", pattern="bw$", full.names=TRUE)
names(tracks) <- gsub("\\.bw","",basename(tracks))

sm_ctcf <- signal2Matrix(tracks, moi_ctcf, w=5, extend=300)

plotEnrichedHeatmaps(sm_ctcf, trim=0.95, multiScale = TRUE, axis_name = "0")
```

The signals of the nucleosome-free fragments shows a very low coverage in the very center and two large peaks adjacent. The low signal in the center corresponds to the region where the TF binds (prevents TN5 insertion) = TF footprint. The two adjecent peaks indicate that the chromatin is accessbile there.  
When looking at the signals of the mono-nucleosome fragments, we see two peaks adjecent to the motifs center. This corresponds to the nucleosomes that get pushed to the side, when the TF binds.  

### ATAC signals around SP1 signals
```{r, fig.width=8, fig.height=4}
tracks <- list.files("data/", pattern="bw$", full.names=TRUE)
names(tracks) <- gsub("\\.bw","",basename(tracks))

sm_sp1 <- signal2Matrix(tracks, moi_sp1, w=5, extend=300)

plotEnrichedHeatmaps(sm_sp1, trim=0.95, multiScale = TRUE, axis_name = "0")
```

The signals of the nucleosome-free fragments show low coverage in the center, probably corresponding to the SP1 binding site.
Interestingly, the signal for the nucleosome-free fragments shows a gradual, continuous increase towards the side (unlike CTCF, which displayed two sharp peaks adjecent to the center). 
For the mono-nucleosome fragments, we also see a rather gradual, continuous increase outwards. This indicates that SP1 has a weaker effect on nucleosome positioning and that the TF has a less pronounced footprint (compared to CTCF).

# Part 2
To identify motifs that are differentially acessible in between males and females in stressed vs control we do the following:  
1. Annotate peaks with GC bias  
2. Get a list of motifs  
3. Scan the ATAC seq peaks for motif matches  
4. Calculate the motif accessability deviations  
5. Perform differential analysis

## Downloading ATAC data from stressed brain
I decided to work with the whole data (stressed and controlled). 
```{r, eval=FALSE}
download.file("https://ethz-ins.org/content/mouse_mm38_hippocampus.peakCounts.SE.rds", "data/peakCounts.rds", mode="wb")
```

```{r}
se <- readRDS("data/peakCounts.rds")
```

## Annotate regions
```{r annotation}
se <- GenomeInfoDb::keepStandardChromosomes(se, pruning.mode = "coarse")
seqlevelsStyle(se) <- "UCSC"
se <- addGCBias(se, genome=BSgenome.Mmusculus.UCSC.mm10)
```

## Motif matches
I will use the motif list from MotifDb. 
```{r motif-list}
motifs <- MotifDb::query(MotifDb, "HOCOMOCOv13")
```

Then we convert the motifs that they get accepted from motifmatchr.

```{r, message=FALSE}
motifs <- do.call(TFBSTools::PWMatrixList, convert_motifs(motifs, class="TFBSTools-PWMatrix"))
names(motifs) <- make.unique(sapply(motifs, \(x) gsub("\\..+","",x@name)))
```

And now we will scan the ATAC peaks for motif matches.

```{r motif-matches}
matches <- motifmatchr::matchMotifs(motifs, se, genome=BSgenome.Mmusculus.UCSC.mm10)
matches
```

## Get motif deviations
Now we will get the motif deviations, showing the realtive accessibility of each motif in each sample.

```{r motif-deviations}
dev <- betterChromVAR(se, matches)
dev
```

## Differential Analysis
As i used the complete dataset (containing both stressed and control) we should ideally account for condition as a covariate. 

```{r assign-groups}
dev$condition <- ifelse(grepl("FSS", colnames(dev)), "Stressed", "Control")
dev$sex <- ifelse(grepl("female", colnames(dev)), "female", "male")
dev
```

```{r differential-expression}
# fit model with sex as main effect and condition as covariate
design <- model.matrix(~sex + condition, data=as.data.frame(colData(dev)))

fit <- eBayes(lmFit(assay(dev, "z"), design))
res <- as.data.frame(topTable(fit, coef=2, number = Inf))
res$motif <- row.names(res)
ggplot(res, aes(logFC, -log10(adj.P.Val), label=motif)) + geom_text()
```

On the following heatmap we see the motifs that show the largest differences regarding their accessability in males compared to females.


```{r heatmap}
sechm(dev, head(row.names(res)), gaps_at="sex", assayName = "z")
```

