---
title: "Clustering and enrichment analysis"
author: "Pierre-Luc"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(epiwraps)
  library(ggplot2)
  library(rGREAT) # Gene Ontology enrichment among genomic regions
})
```

```{r}
BiocManager::install("rGREAT", force = TRUE)
```

Today we're going to work on ChIP-seq of 2 TFs and 2 histone modifications in a mouse cancer line (all restricted to chr1).

## Download the data

```{r, eval=FALSE}
options(timeout = 6000)
download.file("https://ethz-ins.org/content/clustering_practical.zip", "clustering_practical.zip")
unzip("clustering_practical.zip")
list.files()
```

## Prepare the regions and the tracks

```{r}
tracks <- list.files(pattern="bw$")
regions <- rtracklayer::import.bed("regions.bed")
```

## Plot

```{r}
ese <- signal2Matrix(tracks, regions, extend=2000)
plotEnrichedHeatmaps(ese)
ese2 <- ese[1:1000,]
plotEnrichedHeatmaps(ese2, cluster_rows = TRUE, show_row_dend=TRUE )
```

## Clustering

```{r}
set.seed(123)  # to ensure that it gives the same results everytime
cl <- clusterSignalMatrices(ese, k=4)
table(cl)
head(cl)
length(cl)
nrow(ese)

# to make sure the cluster labels stay associated with the corresponding regions/rows
# even if we manipulate the object, put them inside the rowData of the object:
rowData(ese)$cluster <- cl
head(rowData(ese))
```





Plotting the clusters:

```{r}
#plotEnrichedHeatmaps(ese, row_split="cluster", colors=c("white","darkred"))
plotEnrichedHeatmaps(ese, row_split=rowData(ese)$cluster, 
                     colors=c("white","darkred"))
```



Adding colors:

```{r}
mycolors <- c("1"="red", "2"="blue", "3"="darkgreen", "4"="black")
plotEnrichedHeatmaps(ese, row_split="cluster", mean_color=mycolors, colors=c("white","darkred"))
```


Trying different numbers of clusters:


```{r}
cl2 <- clusterSignalMatrices(ese, k=2:10)
ggplot(cl2$varExplained, aes(k, varExplained)) + geom_line()
```

With more clusters, we explain more of the variance. But at some point we have a plateau. Here we would chose something in between 3-4. We cluster with a too large number of clusters - then we scale down the number of clusers (unfortunately very hard to do it automatically). 


Plotting just the averages:

```{r}
d <- meltSignals(ese, splitBy=cl)
ggplot(d, aes(position, mean, colour=sample)) + geom_line(size=1.2) + facet_wrap(~split)
```


Difference of cluster 1 and 4: difference in magnitude of signal could also be due to technical artifacts. So what we do: normalize the rows before clustering.

We can also perform enrichment analysis (before normalization) and see whether the cluster 1 and 4 actually differ frome ach other -> if not, probably due to technical artifacts. And then we go for the clusters obtained with normalization. 
Normalization: we normalize the rows (max value of one row is set to the max value of another row)

Clustering using relative signal instead:

```{r}
set.seed(123)
cl <- clusterSignalMatrices(ese, k=4, scaleRows = TRUE)
d <- meltSignals(ese, splitBy=cl)
ggplot(d, aes(position, mean, colour=sample)) + geom_line() + facet_wrap(~split)
plotEnrichedHeatmaps(ese, row_split = cl, scale_rows = "global")
```


## Enrichment analysis

Some regions in the genome are bound by MAFK + are inactive. 
Some regions are bount by p300 and acethylated etc...
So now we can characterize these regions ->
what kind of genes are near by? 
what CHIP-signals are near by?

Find what's enriched in one cluster with respect to the others:
First we are goung to take the genomic regions and split them by clusters. So the split_regions contains 4 lists with the genomic regions per cluster.

And then we can see which clusters show increased in genes that are in certain functional categories. E.g cluster 1 shows twice as much gene regions that are important for e.g response to unfolded protein.

First question: 
Gene ontology annotates genes with functional categories
- biological processes
- molecular processes
- cellular components

What GEAT does: 
starts with gene regulating domain of a gene, then considers the gene body to be part of regulatory domain. Then it will extend the region until it will encounter another reglatory element (of another gene).
So it assignes regions to genes in a weighted fassion -> binding/peak that is present in the regulatory domain is more important, than binding happening far away. 

Here we compare cluster 3 compared to the others? 

```{r}
# we first split the regions by cluster:
split_regions <- split(rowRanges(ese), rowData(ese)$cluster)
lengths(split_regions)

res <- great(split_regions[["3"]], gene_sets="GO:BP", tss_source="mm10", 
             background=regions, cores=2)
bp <- getEnrichmentTables(res)
head(bp)

```

We plot the top Biological Processes:

```{r, fig.width=9, fig.height=6}
ggplot(head(bp,15), aes(fold_enrichment, reorder(description, p_adjust), 
                        size=observed_region_hits, color=-log10(p_adjust))) + 
  geom_point() + scale_color_viridis_c()
```

3 TF (Creb1, Creb3 and Creb3L1). Look similar when looking at the 
