---
title: "Assignment - Week 06"
author: "Yara Roth"
date: "2026-04-01"
output: html_document
---

# Setup
```{r setup}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(GenomicRanges)
  library(ggplot2)
  library(memes)
  library(motifmatchr)
  library(Biostrings)
  library(MotifDb)
  library(TFBSTools)
  library(universalmotif)
  library(PWMEnrich)
  library(GenomeInfoDb)
})


library(rstudioapi)
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
dir.create("data")
```

# Download
## Chip-seq data
I decided to work with Chip-seq data for the TF REST from mouse myocytes.

```{r download-chipseq}
if(!file.exists("data/myocytes_REST_ENCFF368VWJ.bed.gz")) {
  download.file("https://www.encodeproject.org/files/ENCFF368VWJ/@@download/ENCFF368VWJ.bed.gz", "data/myocytes_REST_ENCFF368VWJ.bed.gz")
}

peaks <- rtracklayer::import("data/myocytes_REST_ENCFF368VWJ.bed.gz", format="NarrowPeak")

seqlevelsStyle(peaks) <- "Ensembl"
```

## Genome sequence
```{r download-genome}
ah <- AnnotationHub()

query_results <- AnnotationHub::query(ah, c("Mus Musculus","TwoBitFile","101","GRCm38","primary_assembly"))
head(query_results)

genome_m <- ah[["AH84785"]]
genome_seqs <- import(genome_m)
```

# Question 1:
## Motif scanning
### Getting the desired motif
We will select the motif for Mus musculus.
```{r motif-identification}
motifs <- query(MotifDb, c("REST"))
names(motifs)
motif <- motifs[["Mmusculus-HOCOMOCOv10-REST_MOUSE.H10MO.A"]]
view_motifs(motifs[1:2])
```

Next we will scan our peaks for motifs, meaning we will go through the peaks and check whether they contain REST motif matches.

```{r motif-scanning-peaks}
Biostrings::writeXStringSet(genome_seqs, "data/genome.fa")
motif2 <- convert_motifs(motif, class="TFBSTools-PWMatrix")
moi <- motifmatchr::matchMotifs(motif2, subject=peaks, genome=Rsamtools::FaFile("data/genome.fa"), out="positions")

moi <- moi[[1]]
head(moi)
hist(moi$score)
```

`moi` is an object that contains every genomic position where a REST motif match was found within a peak.
Meaning we can now look at the percentage of peaks that contain an overlap with a REST motif.

```{r peaks-motif-proportion}
overlap <- table(overlapsAny(peaks, moi))
overlap
overlap_true <- overlap["TRUE"]

peak_count <- length(peaks)
peak_motif_proportion <- 100 / peak_count * overlap["TRUE"]
peak_motif_proportion
```

**ANSWER QUESTION 1**  
Of the `r peak_count` peaks, `r overlap_true` (`r round(peak_motif_proportion,2)`%) contain a motif.

# Question 2:
## Scanning the whole genome

Now we will scan the whole genome for REST motif matches.

```{r motif-scanning-genome}
motif_across_genome <- matchMotifs(motif2, subject=genome_seqs, out="positions")[[1]]
names(motif_across_genome) <- names(genome_seqs)
motif_across_genome <- as(motif_across_genome, "GRanges")
head(motif_across_genome)
```

Now can now look at the percentage of motifs that are occupied by a the transcription factor.

```{r motif-peak-proportion}
overlap_motif <- table(overlapsAny(motif_across_genome, peaks))
overlap_motif
overlap_motif_true <- overlap_motif["TRUE"]

motif_count <- length(motif_across_genome)
motif_peak_proportion <- 100 / motif_count * overlap_motif["TRUE"]
motif_peak_proportion
```

**ANSWER QUESTION 2**  
Of the `r motif_count` motif sites, `r overlap_motif_true` (`r round(motif_peak_proportion, 2)`%) overlap a peak.

# This was only for me:
## Enrichment Analysis
### Preparing sequences
```{r preparing-sequences}
quantile(width(peaks))
peak_centers <- resize(peaks, fix="center", width=100)

peak_seqs <- memes::get_sequence(peak_centers, genome_m)
peak_seqs
as.character(peak_seqs[1])
```

And now we will perform the actual enrichment analysis, meaning we will analyze which motifs are statisically enriched (appear more in peaks than random background).

```{r enrichment-analysis}
motifs <- query(MotifDb, c("Mmusculus", "HOCOMOCOv10"))
Sys.setenv(MEME_BIN = "/Users/yararoth/miniconda3/envs/epigen/bin")
ame <- memes::runAme(peak_seqs, database=convert_motifs(motifs))
head(ame)
```

And now we see that the most enriched motif is REST, which confirms that the REST transcription factor actually recognizes and binds to this motif.
